\textls[-0]{\thetitle}

### 2023-04-02 metatheory-condensed.tex
Mention J in the text.

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
We use based path induction
as the elimination principle for identity types~\citepalias{program2013homotopy}.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
We use $x$ and $\mathit{self}$ to denote variables.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
Definitional equalities have the forms %(for each judgement)
\goodCtx{\Gm_1 \equiv \Gm_2}{},
\goodType{\Gm}{T_1 \equiv T_2}{},
\goodTerm{\Gm}{t_1 \equiv t_2}{T},
etc.
Following \citet{altkap2016}, we regard our (intrinsically typed) syntax
as being quotiented by these equalities.
Quotienting facilitates coercion along equalities---given
\goodCtx{\Gm_1 \equiv \Gm_2}{} and
\goodType{\Gm_1}{T_1 \equiv T_2}{},
the derivation of \goodTerm{\Gm_1}{t}{T_1} is considered definitionally equal to
a derivation of \goodTerm{\Gm_2}{t}{T_2}. %\YZ{Is this accurate??}\EDJreply{Yes. In QIIT model, it looks like `t : Tm G1 T1' and  `t : Tm G2 T2'. In extenional Agda, because ``G1 = G2'', ``T1 = T2'', thus the two types `Tm G2 T2 = Tm G1 T1'. So they are the same derivation in QIIT. In our setup, we have the same idea, because after quotient, G1 refers to same object as G2, so the derivation is the same.}
%\YZ{should there be a convertibility rule?}\EDJreply{in this style, the quotient by judgemental equality is applied, so A is equal to B even in meta-logic, which means t : A is the same as t : B. Maybe this is something require emphasis? Because it is usually handled by LF (like QIIT) but now we are using bare hands so.... maybe we need to emphasize this}
%For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}\EDJreply{(G |- t : T), which looks like (t : Tm G T) in QIIT/Agda is the derivation; so basically the term is the derivation (or proof), the types (Tm G T) is the judgement. It applies to each kind of judgement, for example, (G |- ), which looks like (G : Con), (G |-) is the derivation and (_ |- )is the judgement \\ 
%Note that without referring to QIIT/Agda, the definition of judgement and derivation is still clear -- (G |- _ : T) is a judgement and a concrete proof/derivation we get for t like (G |- t : T) is the derivation. Similar for ( _ |- ) vs. (G |-) and (G |- _ ) vs (G |- T). \\ 
%I think in intrinsic typing, knowing t and knowing (G |- t : T) is exactly the same (i.e. there are no two different derivations lead to the same t; for each t, there is only one derivation for it). This is not true for extrinsic typing. \\
%If this prequisite knowledge is too involved, feel free to delete this sentence, I think I just want to say there are equality/quotient everywhere between all kinds of derivations(proof?data?) -- equality between terms, types, substitutions, contexts, linkage signatures ...\\
%I think this is something standard from Martin Lof, since Martin Lof is the first guy talk about judgement.
%}

@@

\emph{Linkage signatures} have the form $\goodSig{\Gm}{\lsig}{n}$.
As rules \ruleref{lsig/empty} and \ruleref{lsig/add} show, a linkage
signature is a list of $n$~types.
%
\emph{Linkages} have the form $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$,
where $\TyLkg{\lsig}$ is a type formed by~$\lsig$.
As rules \ruleref{l/empty} and \ruleref{l/add} show, a linkage is a list
of $n$~terms, each representing a field of the family modeled by the linkage.

In rules \ruleref{l/add} and \ruleref{lsig/add},
the second premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{\mathit{self}.t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

%recorded by the first premise $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$
The type~$A$ in \ruleref{l/add} and \ruleref{lsig/add},
abstracting the types of the prior fields, does not necessarily contain the same types as those recorded by~$\TyLkg{\lsig}$,
%\EDJ{Suggest changing to ``not necessarily the same as those recorded by lsig in the prior fields''},
because a field defined as the code $\wcode{\wsig}$ of a W-type has to
expose different types to different fields that come after it.
Later fields that invoke
$\Wsup{}{\wsig}{\cDot}{\cDot}$ or $\Wrec{\wsig}{\cDot}{\cDot}$
should see the concrete signature~$\wsig$, as rules \ruleref{tm/wsup}
and \ruleref{tm/wrec} stipulate.
By contrast, $\wsig$ should be hidden
from all other fields, so that they can be reused
in a different context where the W-type signature~$\wsig$ is replaced by an extended one~$\wsig'$.

@@

In rules \ruleref{l/add} and \ruleref{lsig/add},
the third premise $\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$ is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ packages $\lsig$---recall that $\lsig$ contains the
(self-parameterized) types of all the fields preceding the current
field~$t$---into a dependent tuple type (\ruleref{tyeq/pk/add}).
The term $\seal$ turns a tuple of type $\TyTkg{\lsig}$ into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.\footnote{%
In the presence of hiding, it is important that the context type~$A$ be a
dependent tuple type rather than a linkage type; we expand on this point in an
appendix.}

>>>>

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
We use based path induction $\Jeq{\cdot}{r}$
as the elimination principle for 
a term $r$ of identity types~\citepalias{program2013homotopy}. 
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
We use $x$ and $\mathit{self}$ to denote variables.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
Definitional equalities have the forms %(for each judgement)
\goodCtx{\Gm_1 \equiv \Gm_2}{},
\goodType{\Gm}{T_1 \equiv T_2}{},
\goodTerm{\Gm}{t_1 \equiv t_2}{T},
etc.
Following \citet{altkap2016}, we regard our (intrinsically typed) syntax
as being quotiented by these equalities.
Quotienting facilitates coercion along equalities---given
\goodCtx{\Gm_1 \equiv \Gm_2}{} and
\goodType{\Gm_1}{T_1 \equiv T_2}{},
the derivation of \goodTerm{\Gm_1}{t}{T_1} is considered definitionally equal to
a derivation of \goodTerm{\Gm_2}{t}{T_2}. %\YZ{Is this accurate??}\EDJreply{Yes. In QIIT model, it looks like `t : Tm G1 T1' and  `t : Tm G2 T2'. In extenional Agda, because ``G1 = G2'', ``T1 = T2'', thus the two types `Tm G2 T2 = Tm G1 T1'. So they are the same derivation in QIIT. In our setup, we have the same idea, because after quotient, G1 refers to same object as G2, so the derivation is the same.}
%\YZ{should there be a convertibility rule?}\EDJreply{in this style, the quotient by judgemental equality is applied, so A is equal to B even in meta-logic, which means t : A is the same as t : B. Maybe this is something require emphasis? Because it is usually handled by LF (like QIIT) but now we are using bare hands so.... maybe we need to emphasize this}
%For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}\EDJreply{(G |- t : T), which looks like (t : Tm G T) in QIIT/Agda is the derivation; so basically the term is the derivation (or proof), the types (Tm G T) is the judgement. It applies to each kind of judgement, for example, (G |- ), which looks like (G : Con), (G |-) is the derivation and (_ |- )is the judgement \\ 
%Note that without referring to QIIT/Agda, the definition of judgement and derivation is still clear -- (G |- _ : T) is a judgement and a concrete proof/derivation we get for t like (G |- t : T) is the derivation. Similar for ( _ |- ) vs. (G |-) and (G |- _ ) vs (G |- T). \\ 
%I think in intrinsic typing, knowing t and knowing (G |- t : T) is exactly the same (i.e. there are no two different derivations lead to the same t; for each t, there is only one derivation for it). This is not true for extrinsic typing. \\
%If this prequisite knowledge is too involved, feel free to delete this sentence, I think I just want to say there are equality/quotient everywhere between all kinds of derivations(proof?data?) -- equality between terms, types, substitutions, contexts, linkage signatures ...\\
%I think this is something standard from Martin Lof, since Martin Lof is the first guy talk about judgement.
%}

@@

\emph{Linkage signatures} have the judgment form $\goodSig{\Gm}{\lsig}{n}$.
A linkage signature is a list of $n$~types
(\ruleref{lsig/empty} and \ruleref{lsig/add}).
%
\emph{Linkages} have the judgment form $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$,
where $\TyLkg{\lsig}$ is a type formed by~$\lsig$.
A linkage is a list of $n$~terms, each representing a field of the family
modeled by the linkage (\ruleref{l/empty} and \ruleref{l/add}).

In the rules \ruleref{l/add} and \ruleref{lsig/add},
the second premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
%
As discussed later, the third premise $\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$
is responsible for creating the context type~$A$ that possibly hides W-type
signatures in $\lsig$, which records the types of the prior fields.

Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{\mathit{self}.t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

%recorded by the first premise $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$
The type~$A$ in \ruleref{l/add} and \ruleref{lsig/add},
abstracting the types of the prior fields, does not necessarily contain the same types as those recorded by~$\TyLkg{\lsig}$,
%\EDJ{Suggest changing to ``not necessarily the same as those recorded by lsig in the prior fields''},
because a field defined as the code $\wcode{\wsig}$ of a W-type has to
expose different types to different fields that come after it.
Later fields that invoke
$\Wsup{}{\wsig}{\cDot}{\cDot}$ or $\Wrec{\wsig}{\cDot}{\cDot}$
should see the concrete signature~$\wsig$, as the rules \ruleref{tm/wsup}
and \ruleref{tm/wrec} stipulate.
By contrast, $\wsig$ should be hidden
from all other fields, so that they can be reused
in a different context where the W-type signature~$\wsig$ is replaced by an extended one~$\wsig'$.

@@

In \ruleref{l/add} and \ruleref{lsig/add},
the third premise $\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$ is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ packages $\lsig$---recall that $\lsig$ contains the
(self-parameterized) types of all the fields preceding the current
field~$t$---into a dependent tuple type (\ruleref{tyeq/pk/add}).
The term $\seal$ turns a tuple of type $\TyTkg{\lsig}$ into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.\footnote{%
In the presence of hiding, it is important that the context type~$A$ be a
dependent tuple type rather than a linkage type; we expand on this point in an
appendix.}
